---
title: "Changes to the K12 Funding Formulas"
output: beamer_presentation
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Warning

Slides are not done.

## Funding Basics

Details on the interaction of property tax with the state allocation.

+ Get more property tax, get less from the state
+ A few districts in a few years get nothing from the state because high property values and few students

## The Two Major Funding Formulas

From  ORS 327.013

+ the cost sharing transportation formula
+ the non-cost sharing general formula


## How Equal Funding Results in Unequal Class Size and School Year

+ Funding formula gives a roughly fixed dollar amount per student.
+ Districts primarily trade-off class size with school year.
+ Districts with low teacher salaries can afford more teachers per student, or longer school years.
+ Quick illustrative average class size
    + PPS 23(They have extra funding)
    + Parkrose 27
    + David Douglas 26
    + Centennial 27
    + Reynolds 25
+ Rural
    + Bandon 20
    + Baker 12
    + Gervis 21
    + Many in the 10s


```{r ReadInData, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
library(readr)
library(dplyr)

OSBASalary <- read_csv("OSBASalarySurvey/EditedSalarySurvey.csv")

Schools16 <- read.csv("http://www.ode.state.or.us/data/reportcard/media/17/RCmediaSchoolsAggregate.csv")
Schools15 <- read.csv("http://www.ode.state.or.us/data/reportcard/media/16/RCmediaSchoolsAggregate.csv")
Schools14 <- read.csv("http://www.ode.state.or.us/data/reportcard/media/15/RCmediaSchoolsAggregate.csv")

Schools14 %>%
  select(District.ID,
         District.Name,
         School.ID,
         School.Name,
         School.Type,
         Avg.Class.Size) %>%
  mutate( Year = 2014,
          SSClassSize = NA,
          SciClassSize = NA,
          MathClassSize = NA,
          ELAClassSize = NA) -> Schools14

Schools15 %>%
  select(District.ID,
         District.Name,
         School.ID,
         School.Name,
         School.Type,
         Avg.Class.Size = School.Self.Contained.Class.Size,
         SSClassSize = School.Social.Studies.Class.Size,
         SciClassSize = School.Science.Class.Size,
         MathClassSize = School.Math.Class.Size,
         ELAClassSize = School.ELA.Class.Size
         ) %>%
  mutate( Year = 2015,
          Avg.Class.Size = as.numeric(as.character(Avg.Class.Size))) -> Schools15

Schools16 %>%
  select(District.ID,
         District.Name,
         School.ID,
         School.Name,
         School.Type,
         Avg.Class.Size = School.Self.Contained.Class.Size,
         SSClassSize = School.Social.Studies.Class.Size,
         SciClassSize = School.Science.Class.Size,
         MathClassSize = School.Math.Class.Size,
         ELAClassSize = School.ELA.Class.Size
         ) %>%
  mutate( Year = 2016,
          Avg.Class.Size = as.numeric(as.character(Avg.Class.Size))
          ) -> Schools16

Schools <- bind_rows( Schools14, Schools15, Schools16)
rm( Schools14, Schools15, Schools16)

Schools %>%
  left_join(OSBASalary, by = c( "District.ID"= "DistrictID","Year" = "EndYear")) %>%
  mutate(SSClassSize = as.numeric(as.character(SSClassSize)),
         SciClassSize = as.numeric(as.character(SSClassSize)),
         MathClassSize = as.numeric(as.character(SSClassSize)),
         ELAClassSize = as.numeric(as.character(SSClassSize)),
         StudentDays = as.numeric(as.character(StudentDays))
         )-> SchoolsClassSize
rm(Schools,OSBASalary)

```

## Class size, School year and Salaries

+ OSBA does a salary survey every year with a high response rate.
    + Covers salary schedules,
    + Student contact days
+ ODE Reports class sizes and other figures
    + Average class size for Elementary
    + By class type, e.g., for middle and high school.

## Elementary Class Size by Salary (2015)

```{r SalaryClass, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

library(ggplot2)

SchoolsClassSize %>%
  filter(School.Type == "E", Avg.Class.Size >= 10, StudentDays > 160, Year == 2015, Avg.Class.Size < 50) %>%
  ggplot(aes(y = Avg.Class.Size, x = AvgSalary )) +
  geom_point() +
  ylab("Class Size") +
  xlab("Teacher Salary") + geom_smooth()


```

## What it Means

+ These are school level averages.  
+ There are significant differences between schools within districts and within schools
+ Tendency for districts with higher salaries to have higher class sizes.

## Within District Class Size

```{r ClassSizeVar, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

library(ggplot2)

# Find districts with more than 3 elementary schools
# 

SchoolsClassSize %>%
  filter(School.Type == "E",Year == 2015) %>%
  group_by(District.ID) %>%
  summarise(ESchools = n()) %>%
  filter( ESchools > 3)%>% 
  select(District.ID) -> MoreThan3E
           

SchoolsClassSize %>%
  filter(School.Type == "E", Year == 2015) %>%
  semi_join( MoreThan3E) %>%
  select(District.ID, Avg.Class.Size) %>%
  ggplot(aes(y=Avg.Class.Size, x =as.factor(District.ID))) +
  geom_boxplot() + ylim(10,40) + xlab("School Districts with at least Three Elementary Schools")
```

## What about School Year?

```{r SalaryYear, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

library(ggplot2)

SchoolsClassSize %>%
  filter(Year == 2015, School.Type =="E")%>%
  select(District.ID, StudentDays, AvgSalary)%>%
  unique() %>%
  ggplot(aes(y = StudentDays, x = AvgSalary )) +
  geom_point() +
  ylab("Student Days") +
  xlab("Teacher Salary")  + ylim(120,200)
```


## Salary and School  Year

+ No real relationship
+ Some district choose a short school year, others choose a longer one.
+ A small number of days cut often subsitutes for larger class size


## School days and Class Size Together

+ Define Teacher Contact Days
    + $\frac{1}{Class~Size} School~Year$
+ Basic logic
    + In a class of 20, each student gets, $\frac{1}{20}$ of a day of attention.
    + 200 day long school year gives the student, $\frac{1}{20} 200 = 10$ contact days with the teacher.

## Teacher Contact Days vs Salary

```{r SalaryContact, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

library(ggplot2)

SchoolsClassSize %>%
  filter(School.Type == "E", Avg.Class.Size >= 10, StudentDays > 140, Year == 2015, Avg.Class.Size < 50) %>%
  ggplot(aes(y = StudentDays * 1/Avg.Class.Size, x = AvgSalary )) +
  geom_point() +
  ylab("Teacher Contact Days") +
  xlab("Teacher Salary") + geom_smooth() + ylim(4,10)


```

## What is Going 

+ There are some key disticts with average salaries in the 55k to 63K salary range that are keeping some schools with small class sizes.
+ That pops up the Teacher contact days in that range when you look per school.
+ Does not mean it is true per student.

## Teacher Contact Days By Student

## But What About the QEM?  

+ Even if you fund the Quality Education Model fully
+ High cost areas will still have high class sizes, lower test scores and lower graduation rates than the low cost areas.
+ Funding that puts class size in high cost areas into the low 20s would put class size in rural, low cost, areas into the low teens or even single digits.
+ Equal funding is not the same as equitable treatment.

## Could you move funding to increase aggregate graduation rates?

+ Decreasing class size when class sizes are large generates larger increases than decreasing class sizes when class size is small.

+ IOW, Moving from 27 to 26 increases graduation rates more than decreasing class size from 20 to 19.

    
## Can we Change the funding formula?

It depends ...

Oregon Constitution VIII.8(2)

 (2) Consistent with such legal obligation as it may have to maintain substantial equity in state funding ...
 
 + We have variations in the transportation component by cost
 + We have variations in the general component based on experience.
 
##  What is the change?

$$F * (\$4,500 + \$25(Average~Teacher~Experience - State~Average)) * ADMw$$

to

$$F * (\$4,500 + m * Average~Annual~Teacher~Salary) * ADMw$$

The key is m.

+ F is the "Funding Percentage" which adjusts each district's funding to meet the state budget allocation.
+ Big m give more equal class size but reduces incentives to negotiate over salary
+ Small m gives less equal class size but increases incentives to negotiate

## The politics

+ Rural vs Urban
    + Rural will object but:
        + Transportation takes into account cost difference why not the general
        + They often get more money because of the small schools grant (another nod to cost difference)
        
## Supporting Data 