---
title: "Converting Oregon's K12 Funding Formula for More Equal Class Size and Higher Aggregate Graduation Rates"
short: "Cost Sharing"
journal: "AER" # AER, AEJ, PP, JEL
month: "`r lubridate::month(Sys.time())`"
year: "`r lubridate::year(Sys.time())`"
vol: 1
issue: 1
jel:
  - A10
  - A11
keywords:
  - K12 Education
  - Cost Sharing Contracts
author:
  - name: James Woods
    firstname: James
    surname: Woods
    email: woodsj@pdx.edu
    affiliation: Portland State University

acknowledgements: |
  ""
abstract: |
  Oregon Equalized funding per student as part of its property tax reform in the 1990s.  The funding formula makes some adjustments for the kinds of students, e.g., pregnant or low-income, to  produce weighted average daily measurement. It then adjusts for local property tax collections and makes a small correction  for more experienced faculty.  Because the forumula does not take into account salary differences, class size varies considerably accross school districts.
  This paper shows that there are decreasing returns from reducing class sizes with respect to graduation rates, i.e.,reducing larger classes results is larger increases in the graduation rate and that the funding formula can be adapted to reduce variance in class size while increasing the aggregate graduation rate in the state without increases in funding. 
output: rticles::aea_article
---



```{r LoadMembershipData, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}

# Enrollment Reports
# 

library(readxl)


# Read in from local.  Error in http read so resorted to this.

Membership10 <- read_excel("../MembershipReport/fallmembershipreport_20092010.xls", sheet = "School")

Membership11 <- read_excel("../MembershipReport/fallmembershipreport_20102011.xls", sheet = "School (Final)")

Membership12 <- read_excel("../MembershipReport/fallmembershipreport_20112012.xls", sheet = "School (Final)")

Membership13 <- read_excel("../MembershipReport/fallmembershipreport_20122013.xls", sheet = "School (Final)")

Membership14 <- read_excel("../MembershipReport/fallmembershipreport_20132014.xls", sheet = "School (1314)")

Membership15 <- read_excel("../MembershipReport/fallmembershipreport_20142015.xls", sheet = "School (1415)")

Membership16 <- read_excel("../MembershipReport/fallmembershipreport_20152016.xlsx", sheet = "School (15-16)")

Membership17 <- read_excel("../MembershipReport/fallmembershipreport_20162017.xlsx", sheet = "School (16-17)")

#Add year indicator to each dataframe
#
Membership10$Year <- 2010
Membership11$Year <- 2011
Membership12$Year <- 2012
Membership13$Year <- 2013
Membership14$Year <- 2014
Membership15$Year <- 2015
Membership16$Year <- 2016
Membership17$Year <- 2017

# Create common colum names for  data we need
# 

CommonNames1217 <-  c("DistrictID"              
                      ,"District"                                       
                      ,"SchoolID"                
                      ,"School"                                         
                      ,"PrevEnrollment"                       
                      ,"Enrollment"                       
                      ,"Male"                                  
                      ,"PercentMale"                             
                      ,"Female"                                 
                      ,"PercentFemale"                            
                      ,"AmericanIndianAlaskaNative"         
                      ,"PercentAmericanIndianAlaskaNative"    
                      ,"Asian"                                  
                      ,"PercentAsian"                             
                      ,"NativeHawaiianPacificIslander"      
                      ,"PercentNativeHawaiianPacificIslander"
                      ,"AfricanAmerican"                 
                      ,"PercentAfricanAmerican"           
                      ,"Hispanic"                       
                      ,"PercentHispanic"                  
                      ,"White"                                  
                      ,"PercentWhite"                            
                      ,"Multiracial"                            
                      ,"PercentMultiracial"                       
                      ,"Kindergarten"                           
                      ,"GradeOne"                              
                      ,"GradeTwo"                              
                      ,"GradeThree"                            
                      ,"GradeFour"                             
                      ,"GradeFive"                             
                      ,"GradeSix"                              
                      ,"GradeSeven"                            
                      ,"GradeEight"                            
                      ,"GradeNine"                             
                      ,"GradeTen"                              
                      ,"GradeEleven"                           
                      ,"GradeTwelve"                           
                      ,"Year") 



names(Membership12) <- CommonNames1217
names(Membership13) <- CommonNames1217
names(Membership14) <- CommonNames1217
names(Membership15) <- CommonNames1217
names(Membership16) <- CommonNames1217
names(Membership17) <- CommonNames1217

# Only deal with counts not percents
# 
# 
library(dplyr)
Membership12 %>% select(-contains("Percent")) -> Membership12
Membership13 %>% select(-contains("Percent")) -> Membership13
Membership14 %>% select(-contains("Percent")) -> Membership14
Membership15 %>% select(-contains("Percent")) -> Membership15
Membership16 %>% select(-contains("Percent")) -> Membership16
Membership17 %>% select(-contains("Percent")) -> Membership17

Membership10 %>% select(-contains("%")) -> Membership10
Membership11 %>% select(-contains("%")) -> Membership11



Membership10 %>%
  rename(
    DistrictID = "Attnd\nDistInstID",
    SchoolID = "Attnd\nSchlInstID",  
    Enrollment = "2009-10\nTotal Enrollment",                 
    Male = "2009-10 Male",                                    
    Female ="2009-10\nFemale",                                                      
    AmericanIndianAlaskaNative="2009-10\nAmerican Indian/Alaskan Native\n(Non-Hispanic)",                  
    
    AfricanAmerican= "2009-10\nBlack/African American\n(Non-Hispanic)",
    Hispanic= "2009-10\nHispanic/Latino",                                                 
    White = "2009-10\nWhite\n(Non-Hispanic)",                                           
    Multiracial = "2009-10\nMultiracial\n(Non-Hispanic)",
    
    Kindergarten = "2009-10 Kindergarten",                                                     
    GradeOne = "2009-10 Grade One",                                                        
    GradeTwo = "2009-10 Grade Two",                                                        
    GradeThree = "2009-10 Grade Three",                                                      
    GradeFour = "2009-10 Grade Four",                                                       
    GradeFive = "2009-10 Grade Five",                                                       
    GradeSix = "2009-10 Grade Six",                                                        
    GradeSeven = "2009-10 Grade Seven",                                                      
    GradeEight = "2009-10 Grade Eight",                                                      
    GradeNine = "2009-10 Grade Nine",                                                       
    GradeTen = "2009-10 Grade Ten",                                                        
    GradeEleven = "2009-10 Grade Eleven",                                                     
    GradeTwelve = "2009-10 Grade Twelve",                                                     
    NativeHawaiianPacificIslander = `For Future Reference\n2009-10\nPacific Islander\n(Non-Hispanic)`,
    
    Asian = `For Future Reference\n2009-10\nAsian\n(Non-Hispanic)`
  ) %>%
  select( -contains("2009-10")) -> Membership10

Membership10$PrevEnrollment <- NA

Membership11 %>% 
  rename( DistrictID = "Attending\nDistrict\nInstID",                                                                                       
          SchoolID = "Attending\nSchool\nInstID",                                                     
          PrevEnrollment = "2009-10\nTotal Enrollment",                                                     
          Enrollment = "2010-11\nTotal Enrollment",                                                     
          Male = "2010-11 Male",                                                                  
          Female = "2010-11 Female",                                                                
          AmericanIndianAlaskaNative = "2010-11\nAmerican Indian/Alaskan Native\n(Non-Hispanic)",                       
          
          AfricanAmerican = "2010-11\nBlack/African American\n(Non-Hispanic)",                              
          Hispanic = "2010-11\nHispanic/Latino",                                                      
          White = "2010-11\nWhite\n(Non-Hispanic)",                                                
          Multiracial = "2010-11\nMultiracial\n(Non-Hispanic)",                                          
          Kindergarten = "2010-11\nKindergarten",                                                         
          GradeOne = "2010-11\nGrade One",                                                            
          GradeTwo = "2010-11\nGrade Two",                                                            
          GradeThree = "2010-11\nGrade Three",                                                          
          GradeFour = "2010-11\nGrade Four",                                                           
          GradeFive = "2010-11\nGrade Five",                                                           
          GradeSix = "2010-11\nGrade Six",                                                            
          GradeSeven = "2010-11\nGrade Seven",                                                          
          GradeEight = "2010-11\nGrade Eight",                                                          
          GradeNine = "2010-11\nGrade Nine",                                                           
          GradeTen = "2010-11\nGrade Ten",                                                            
          GradeEleven = "2010-11\nGrade Eleven",                                                         
          GradeTwelve = "2010-11\nGrade Twelve",                                   
          Asian = "For Future Reference\n2010-11\nAsian only\n(Non-Hispanic)", 
          
          NativeHawaiianPacificIslander  = "2010-11\nAsian/Pacific Islander\n(Non-Hispanic)",            
          Year  = "Year" 
  ) %>%   
  select( -contains("2010-11")) -> Membership11

# Enrollment and PrevEnrollment contains notes.  Set those to NA
Membership10$Enrollment <- as.numeric(Membership10$Enrollment)
Membership11$Enrollment <- as.numeric(Membership11$Enrollment)
Membership12$Enrollment <- as.numeric(Membership12$Enrollment)
Membership13$Enrollment <- as.numeric(Membership13$Enrollment)
Membership14$Enrollment <- as.numeric(Membership14$Enrollment)
Membership15$Enrollment <- as.numeric(Membership15$Enrollment)
Membership16$Enrollment <- as.numeric(Membership16$Enrollment)
Membership17$Enrollment <- as.numeric(Membership17$Enrollment)

Membership10$PrevEnrollment <- as.numeric(Membership10$PrevEnrollment)
Membership11$PrevEnrollment <- as.numeric(Membership11$PrevEnrollment)
Membership12$PrevEnrollment <- as.numeric(Membership12$PrevEnrollment)
Membership13$PrevEnrollment <- as.numeric(Membership13$PrevEnrollment)
Membership14$PrevEnrollment <- as.numeric(Membership14$PrevEnrollment)
Membership15$PrevEnrollment <- as.numeric(Membership15$PrevEnrollment)
Membership16$PrevEnrollment <- as.numeric(Membership16$PrevEnrollment)
Membership17$PrevEnrollment <- as.numeric(Membership17$PrevEnrollment)


# Join them
Membership <- bind_rows(Membership10, Membership11, Membership12, Membership13, Membership14, Membership15, Membership16, Membership17) 

# Clean space
remove(Membership10, Membership11, Membership12, Membership13, Membership14, Membership15, Membership16, Membership17, CommonNames1217)

```


```{r LoadTestScores, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}

# Source http://www.ode.state.or.us/data/reportcard/Media.aspx


# Read in

Schools16 <- read.csv("http://www.ode.state.or.us/data/reportcard/media/17/RCmediaSchoolsAggregate.csv")

Schools15 <- read.csv("http://www.ode.state.or.us/data/reportcard/media/16/RCmediaSchoolsAggregate.csv")

Schools14 <- read.csv("http://www.ode.state.or.us/data/reportcard/media/15/RCmediaSchoolsAggregate.csv")

Schools13 <- read.csv("http://www.ode.state.or.us/data/reportcard/media/14/RCmediaSchoolsAggregate.csv")

Schools12 <- read.csv("http://www.ode.state.or.us/data/reportcard/media/13/RCmediaSchoolsAggregate.csv")

Schools11 <- read.csv("http://www.ode.state.or.us/data/reportcard/media/12/RCmediaSchoolsAggregate.csv")

Schools10 <- read.csv("http://www.ode.state.or.us/data/reportcard/media/11/RCmediaSchoolsAggregate.csv")


# Normalize Columns

Schools16 %>%
  mutate(Year = 2016) %>%
  select(DistrictID = District.ID, 
         SchoolID = School.ID, 
         Year,
         Type = School.Type,
         ELA = ELA.Pct.Met.2015.16,
         Math = Math.Pct.Met.2015.16,
         Science = Science.Pct.Met.2015.16,  #New vars follow
         
         PrevGradRate = Four.Year.Graduation.Rate.2014.15,
         ELAClassSize = School.ELA.Class.Size,
         MathClassSize = School.Math.Class.Size,
         SciClassSize = School.Science.Class.Size,
         SSClassSize = School.Social.Studies.Class.Size,
         
         PctELL = Pct.Ever.English.Learners,                
         PctED = Pct.Economically.Disadvantaged,           
         PctDisabled = Pct.Students.with.Disabilities,           
         PctAsian ="Pct.Asian",                                
         PctBlack ="Pct.Black",                                
         PctHispanic = "Pct.Hispanic",                             
         PctNative = "Pct.Native.American",                      
         PctMult = "Pct.Multi.racial",                         
         PctPacifcIsland = "Pct.Pacific.Islander",                     
         PctWhite = "Pct.White",   
         PctMobile = Pct.Mobile.Students.2015.16
         ) -> Schools16

Schools15 %>%
  mutate(Year = 2015) %>%
  select(DistrictID = District.ID, 
         SchoolID = School.ID, 
         Year,
         Type = School.Type,
         ELA = ELA.Pct.Met.2014.15,
         Math = Math.Pct.Met.2014.15,
         Science = Science.Pct.Met.2014.15,
         #New vars follow
         
         GradRate = State.Four.Year.Graduation.Rate,
         ELAClassSize = School.ELA.Class.Size,
         MathClassSize = School.Math.Class.Size,
         SciClassSize = School.Science.Class.Size,
         SSClassSize = School.Social.Studies.Class.Size,
         
         PctELL = Pct.Ever.English.Learners,                
         PctED = Pct.Economically.Disadvantaged,           
         PctDisabled = Pct.Students.with.Disabilities,           
         PctAsian ="Pct.Asian",                                
         PctBlack ="Pct.Black",                                
         PctHispanic = "Pct.Hispanic",                             
         PctNative = "Pct.Native.American",                      
         PctMult = "Pct.Multi.racial",                         
         PctPacifcIsland = "Pct.Pacific.Islander",                     
         PctWhite = "Pct.White",   
         PctMobile = Pct.Mobile.Students.2015.16
         
         ) -> Schools15

Schools14 %>%
  mutate(Year = 2014) %>%
  select(DistrictID = District.ID, 
         SchoolID = School.ID, 
         Year,
         Type = School.Type,
         Reading = Reading.Pct.Met.2013.14,
         Writing = Writing.Pct.Met.2013.14,
         Math = Math.Pct.Met.2013.14,
         Science = Science.Pct.Met.2013.14) -> Schools14

Schools13 %>%
  mutate(Year = 2013) %>%
  select(DistrictID = District.ID, 
         SchoolID = School.ID, 
         Year,
         Type = School.Type,
         Reading = Reading.Pct.Met.2012.13,
         Writing = Writing.Pct.Met.2012.13,
         Math = Math.Pct.Met.2012.13,
         Science = Science.Pct.Met.2012.13) -> Schools13


Schools12 %>%
  mutate(Year = 2012,DistrictID = NA ) %>%
  select(DistrictID, 
         SchoolID = InstID, 
         Year,
         Type = SchoolType,
         Reading = ReadStudExceedMeetAllGradePct1,
         Writing = WriteStudExceedMeetAllGradePct1,
         Math = MathStudExceedMeetAllGradePct1,
         Science = ScienceStudExceedMeetAllGradePct1) -> Schools12


Schools11 %>%
  mutate(Year = 2011, DistrictID = NA ) %>%
  select(DistrictID, 
         SchoolID = InstID, 
         Year,
         Type = SchoolType,
         Reading = ReadStudExceedMeetAllGradePct1,
         Writing = WriteStudExceedMeetAllGradePct1,
         Math = MathStudExceedMeetAllGradePct1,
         Science = ScienceStudExceedMeetAllGradePct1) -> Schools11


Schools10 %>%
  mutate(Year = 2010, DistrictID = NA ) %>%
  select(DistrictID, 
         SchoolID = InstID, 
         Year,
         Type = SchoolType,
         Reading = ReadStudExceedMeetAllGradePct1,
         Writing = WriteStudExceedMeetAllGradePct1,
         Math = MathStudExceedMeetAllGradePct1,
         Science = ScienceStudExceedMeetAllGradePct1) -> Schools10


TestScores <- bind_rows(Schools10, Schools11, Schools12, Schools13, Schools14, Schools15, Schools16) 
rm(Schools10, Schools11, Schools12, Schools13, Schools14, Schools15, Schools16)

# Fix top, bottom and NA coding

TestScores %>%
  mutate(Reading = as.numeric(case_when(
    Reading == ">95"  ~  "95",
    Reading == "<5" ~ "5",
    Reading == "*" ~ "NA",
    Reading == "--" ~ "NA",
    TRUE ~ Reading))) %>%
  mutate(Writing = as.numeric(case_when(
    Writing == ">95"  ~  "95",
    Writing == "<5" ~ "5",
    Writing == "*" ~ "NA",
    Writing == "--" ~ "NA",
    TRUE ~ Writing))) %>%
  mutate(Math = as.numeric(case_when(
    Math == ">95"  ~  "95",
    Math == "<5" ~ "5",
    Math == "*" ~ "NA",
    Math == "--" ~ "NA",
    TRUE ~ Math))) %>%
  mutate(Science = as.numeric(case_when(
    Science == ">95"  ~  "95",
    Science == "<5" ~ "5",
    Science == "*" ~ "NA",
    Science == "--" ~ "NA",
    TRUE ~ Science))) %>%
  mutate(ELA = as.numeric(case_when(
    ELA == ">95"  ~  "95",
    ELA == "<5" ~ "5",
    ELA == "*" ~ "NA",
    ELA == "--" ~ "NA",
    TRUE ~ ELA))) ->TestScores

# Change to numeric  
TestScores %>%
  mutate(Reading = as.numeric(Reading),
         Writing = as.numeric(Writing),
         Math = as.numeric(Math),
         Science = as.numeric(Science),
         ELA = as.numeric(ELA)) ->TestScores

# Fix High school type
# 

TestScores %>%
  mutate(Type = case_when(
    Type == "HSG" ~ "H",
    Type == "HSA" ~ "H",
    TRUE ~ Type)) ->TestScores

#Fill missing district codes

TestScores %>%
  filter(!is.na(DistrictID)) %>%
  select(DistrictID, SchoolID) %>%
  unique() -> DistrictSchoolPairs

TestScores %>%
  left_join(DistrictSchoolPairs, by = "SchoolID", suffix = c(".KILL","")) %>%
  select(-contains(".KILL")) -> TestScores

rm(DistrictSchoolPairs)  
```

```{r LoadFinance, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
DistrictFinance <- read.csv("https://s3-us-west-2.amazonaws.com/woodsjeeschools/ActualExpendituresFundFunction30.CSV") 

library(stringr)
library(dplyr)

# This still has ESD numbers in it and they have odd headcounts on students and expenditures.

Membership %>%
  select(DistrictID, Year, Enrollment) %>%
  group_by(DistrictID, Year) %>%
  summarise(DistrictEnrollment = sum(Enrollment, na.rm = TRUE)) -> DistrictEnrollment

DistrictFinance %>%
  mutate(YearBeginning = as.numeric(str_sub(as.character(SchlYr),-4) )) %>%
  mutate(Year = YearBeginning + 1) %>%
  filter(MajFuncDesc == "Instruction" & Amount > 0 & Year > 2009) %>%
  select(DistInstID, Year, Amount) %>%
  group_by(DistInstID, Year) %>%
  summarise(InstExpenditures = sum(Amount, na.rm = TRUE)) %>%
  left_join(DistrictEnrollment, by = c("DistInstID" = "DistrictID", "Year"), suffix = c("", ".KILL")) %>%
  select( -contains(".KILL")) %>%
  filter(DistrictEnrollment > 0) %>%
  mutate(PerStudentExp = InstExpenditures/DistrictEnrollment) -> DistrictFinance

  
```


American Economic Review Pointers:

\begin{itemize}
\item Do not use an "Introduction" heading. Begin your introductory material
before the first section heading.

\item Avoid style markup (except sparingly for emphasis).

\item Avoid using explicit vertical or horizontal space.

\item Captions are short and go below figures but above tables.

\item The tablenotes or figurenotes environments may be used below tables
or figures, respectively, as demonstrated below.

\item If you have difficulties with the mathtime package, adjust the package
options appropriately for your platform. If you can't get it to work, just
remove the package or see our technical support document online (please
refer to the author instructions).

\item If you are using an appendix, it goes last, after the bibliography.
Use regular section headings to make the appendix headings.

\item If you are not using an appendix, you may delete the appendix command
and sample appendix section heading.

\item Either the natbib package or the harvard package may be used with bibtex.
To include one of these packages, uncomment the appropriate usepackage command
above. Note: you can't use both packages at once or compile-time errors will result.

\end{itemize}

\section{First Section in Body}

Sample figure:

\begin{figure}
Figure here.

\caption{Caption for figure below.}
\begin{figurenotes}
Figure notes without optional leadin.
\end{figurenotes}
\begin{figurenotes}[Source]
Figure notes with optional leadin (Source, in this case).
\end{figurenotes}
\end{figure}

Sample table:

\begin{table}
\caption{Caption for table above.}

\begin{tabular}{lll}
& Heading 1 & Heading 2 \\
Row 1 & 1 & 2 \\
Row 2 & 3 & 4%
\end{tabular}
\begin{tablenotes}
Table notes environment without optional leadin.
\end{tablenotes}
\begin{tablenotes}[Source]
Table notes environment with optional leadin (Source, in this case).
\end{tablenotes}
\end{table}

References here (manual or bibTeX). If you are using bibTeX, add your bib file
name in place of BibFile in the bibliography command.
% Remove or comment out the next two lines if you are not using bibtex.
\bibliographystyle{aea}
\bibliography{references}

% The appendix command is issued once, prior to all appendices, if any.
\appendix

\section{Mathematical Appendix}
